---
title: "Stegen 01 Introduction"
author: "Janetta E. Skarp"
date: "4 October 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

On 26 June 1998, the St Sebastian High School in Stegen (school A), Germany, celebrated a graduation party, where 250 to 350 participants were expected. Attendants included graduates from that school, their families and friends, teachers, 12th grade students and some graduates from a nearby school (school B).

A self-service party buffet was supplied by a commercial caterer in Freiburg. Food was prepared the day of the party and transported in a refrigerated van to the school.

Festivities started with a dinner buffet open from 8.30 pm onwards and were followed by a dessert buffet offered from 10 pm. The party and the buffet extended late during the night and alcoholic beverages were quite popular. All agreed it was a party to be remembered.

## The alert
On 2nd July 1998, the Freiburg local health office reported to the Robert Koch Institute (RKI) in Berlin the occurrence of many cases of gastroenteritis following the graduation party described above. More than 100 cases were suspected among participants and some of them were admitted to nearby hospitals. Sick people suffered from fever, nausea, diarrhoea and vomiting lasting for several days. Most believed that the tiramisu consumed at dinner was responsible for their illness. *Salmonella enteritidis* was isolated from 19 stool samples.

The Freiburg health office sent a team to investigate the kitchen of the caterer. Food preparation procedures were reviewed. Food samples, except tiramisu (none was left over), were sent to the laboratory of Freiburg University. Microbiological analyses were performed on samples of the following: brown chocolate mousse, caramel cream, remoulade sauce, yoghurt dill sauce, and 10 raw eggs.

The Freiburg health office requested help from the RKI in the investigation to assess the magnitude of the outbreak and identify potential vehicle(s) and risk factors for transmission in order to better control the outbreak

## The study
Cases were defined as any person who had attended the party at St Sebastian High School who suffered from diarrhoea
(min. 3 loose stool for 24 hours) between 27 June and 29 June 1998; or who suffered from at least three of the following symptoms: vomiting, fever over 38.5° C, nausea, abdominal pain, headache.

Students from both schools attending the party were asked through phone interviews to provide names of persons who
attended the party.

Overall, 291 responded to enquiries and 103 cases were identified.

\newpage
## An introduction to the *R* companion
This text was adapted from the introduction used at the 2016 TSA module.

R packages are bundles of functions which extend the capability of R. Thousands of add-on packages are available in the main online repository (known as CRAN) and many more packages in development can be found on GitHub. They may be installed and updated over the Internet.

We will mainly use packages which come ready installed with R (base code), but where it makes things easier we will use add-on packages. In addition, we have included a few extra functions to simplify the code required. All the R packages you need for the exercises can be installed over the Internet.

## Setting up

### Required R packages and functions

Install and load the required R packages for this practical.

> **n.b.** you should only need to do this once.

```{r stegen_hw_difficult-1, echo = FALSE, eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}
# Installing required packages for the week
required_packages <- c("epiR", "Hmisc", "epitools", "here") 
install.packages(required_packages)
```

```{r stegen_hw_difficult-2, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE}
library("epiR")
library("Hmisc")
library("epitools")
library("here")
```

Run the following code to make sure that you have all the functions that you need for this practical.

```{r stegen_hw_difficult-3}
# Function to make tables with counts, proportions and cumulative sum
big_table <- function(data, useNA = "no") {
  count <- table(data, useNA = useNA)
  prop <- round(prop.table(count)*100, digits = 2)
  cumulative <- cumsum(prop)
  rbind(count,
        prop,
        cumulative) 
}

 # Function to provide counts, denominator and proportions (equivalent of attack rate)
attack_rate <- function(table) {
  prop <- round(prop.table(table, 1), digits = 2)
  denominator <- rowSums(table) 
  output <- cbind(Ill = table[, 2], N = denominator, Proportions = prop[, 2])
  return(output)
}
```

The big_table function uses data directly and allows combining of counts, proportions and cumulative sums, thus reducing the number of lines of code required for descriptive analyses. The attack_rate function makes tables that combine counts, proportions and row sums.

### The dataset

In this practical, we will be using the tirav12.csv file located in the data folder.

Read in the dataset for this practical.

```{r stegen_hw_difficult-4, echo = FALSE, results = 'hide'}
tira_data <- read.csv(here::here("data", "tirav12.csv"), stringsAsFactors = FALSE)
```

# Question 1
### a) Browse your dataset
What variables does your dataset contain?

### b) Recode data
Identify variables with missing values (variables that have a records with a value of 9). Recode these to NA.

### c) Save your recoded dataset
As a csv file.

\newpage
# Hints
## Question 1
### a) Browse your dataset
*RStudio* has the nice feature that everything is in one browser window, so you can browse your dataset and your code without having to switch between browser windows. 

```{r stegen_hw_difficult-5, eval = FALSE}
# to browse your data, use the View command
View(tira_data)
```

Alternatively, you can also view your dataset by clicking on **tira_data** in the top right "global environment" panel of your *RStudio* browser.  Your global environment is where you can see all the datasets, functions and other things you have loaded in the current session. 

### b) Recode data
Use the "describe" command to assess your data and identify variables with missing values. The describe command showed that the variables salmon, pork and horseradish have a few records with a value of 9. These need to be recoded to NA.

* Using the square brackets "[...]" after a variable allows you to subset for certain observations. To recode values of 9 to NA for the pork variable, select observations where pork **(tira_data\$pork)** is equal to 9 **[tira_data\$pork == 9]** and set these observations equal to NA 
    
* Always use the double equals "==" within square brackets; this a logical (Boolean) operator

* Use "! = " when you want to write "not equal to"

```{r stegen_hw_difficult-8, eval = TRUE}
# The first line below is read as follows:  assign a value of NA to tira_data$pork WHERE tira_data$pork is equal to 9
tira_data$pork[tira_data$pork == 9] <- NA

tira_data$salmon[tira_data$salmon == 9] <- NA

tira_data$horseradish[tira_data$horseradish == 9] <- NA
```

